Lab 2 Final Report: COVID-19 Case Rate vs Population Demographics and Mask Policy

### 2020-12-05 Jackson_Liu_Temlock_Zhang_Lab2_Draft_Report.Rmd
Latest

```{r variables, warning=FALSE, results='hide'}
df <- read.csv("covid-19.csv", header = TRUE)
df<-df%>%
  rename(case_rate_100k = 'Case.Rate.per.100000',
         death_rate_100k = 'Death.Rate.per.100000',
         population_density = 'Population.density.per.square.miles',
         white_pct = 'White...of.Cases',
         black_pct = 'Black...of.Cases',
         hispanic_pct = 'Hispanic...of.Cases',
         other_pct = 'Other...of.Cases',
         state_emergency = 'State.of.emergency',
         business_closed = 'Closed.other.non.essential.businesses',
         business_reopen = 'Began.to.reopen.businesses.statewide',
         mask_public='Mandate.face.mask.use.by.all.individuals.in.public.spaces',
         mask_legal='No.legal.enforcement.of.face.mask.mandate',
         black_population_pct ="Black...of.Total.Population",
         white_population_pct = "White...of.Total.Population",
         poverty_pct = "Percent.living.under.the.federal.poverty.line..2018.",
         unemployed_pct = "Percent.Unemployed..2018.",
         senior_pct = "X65.",
         population_2018 = "Population.2018", ## extra variable in test
         non_elderly_pre_existing = "Nonelderly.Adults.Who.Have.A.Pre.Existing.Condition",
         homeless = "Number.Homeless..2019.") %>%
  select(State, case_rate_100k, death_rate_100k,population_density, white_pct, black_pct, hispanic_pct, 
         other_pct, state_emergency, business_closed, business_reopen, mask_public, mask_legal, 
         black_population_pct, white_population_pct, poverty_pct, unemployed_pct, senior_pct,
         population_2018, non_elderly_pre_existing, homeless)
# Assign correct data types
cols.num <- c("white_pct","black_pct","hispanic_pct","other_pct")
df[cols.num] <- sapply(df[cols.num],as.numeric)
df$state_emergency=as.Date(df$state_emergency, format = "%m/%d/%Y")
df$business_closed=as.Date(df$business_closed, format = "%m/%d/%Y")
df$business_reopen=as.Date(df$business_reopen, format = "%m/%d/%Y")
head(df)
```

## 1. An Introduction

@@

### Assumptions

@@

## 2. A Model Building Process

@@

### Model Variables

@@

In order to allow for more simplistic and direct interpretations when analyzing the effect of the input variables on the output, it was determined that all percentage variables (senior, poverty, and unemployment) would be transformed into rates per 100,000 to align with the case rate variable. It will also serve to ease the burden on the readers of this report as like-for-like scaled relationships can be drawn between the variables. Given that this transformation is constant for all values (scalar multiplication of 100,000 for the seniors variable, and 1,000 for poverty and unemployment), it will have no effect on the distributions of the variables.

```{r}  
# Percentage variables are re-scaled to rates per 100k

df <-df %>%
  mutate(    
    senior_per_100k = 100000*senior_pct,
    poverty_per_100k = 100000*poverty_pct/100, 
    unemployed_per_100k = 100000*unemployed_pct/100,
  )
```

### Model 1

@@

#### Model 1 Regression

@@

#### Model 1 CLM Assumptions

##### i. IID Sampling

@@

##### ii. Linear Conditional Expectation

@@

##### iii. No Perfect Collinearity

@@

##### iv. Homoskedastic Errors

@@

##### v. Normally Distributed Errors

@@

#### Influence of Data Points in Model 1

@@

### Model 2

Based on the work of model 1, two extra state demographic information regarding poverty rate (percentage living under the federal poverty line in 2018) and unemployment rate (percentage of unemployed in 2018) would be included in the analysis of model 2. In addition, both newly-introduced percentage variables would be transformed into rates per 100,000 to align with the case rate variable.

@@

```{r}
ggplot(data = df,
  mapping = aes(x= poverty_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(6000,21200,800)) +
  labs(title =  "Histogram of Poverty Rate", 
       x = "Number of living under poverty line per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(6000, 21200, 1600))
```
As shown in the above distribution, there is a fairly normal distribution in the poverty number 
per 100,000 across the states. Additionally, there is no outliers observed within the distribution,
given that the range of rates fall between 7,600 and 19,700. Therefore, there is no further transformation in need for this variable in the following analysis process for both model 2 and model 3.

@@

```{r}
ggplot(data = df,
  mapping = aes(x= unemployed_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(2000,8000,500)) +
  labs(title =  "Histogram of Unemployment Rate", 
       x = "Number of unemployed per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(2000, 8000, 1000))
```
According to the histogram above, the distribution of unemployed number per 100,000 is of fairly normal distribution. Overall, it is not highly skewed or heavily tailed. Therefore, there is no further transformation in need for this variable in the following analysis process for both model 2 and model 3.

#### Model 2 Regression

@@

```{r message=FALSE}
case_poverty <- df %>% 
  ggplot(aes(poverty_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = ' Association between Covid-19 ase rate and the poverty rate', 
    x = 'Number of living under poverty line per 100,000',
    y = 'Covid-19 care rate per 100,000', 
    color = 'Senior num per 100k'
  )

case_unemployed <- df %>% 
  ggplot(aes(unemployed_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Association between Covid-19 ase rate and the unemployment rate', 
    x = 'Number of unemployed per 100,000', 
    y = 'Covid-19 care rate per 100,000',
    color = 'Senior num per 100k'
  )

case_poverty / case_unemployed
```
However, there is an overall increasing trend of COVID-19 care rate with increasing poverty rate, while there is a inverse relationship between the variables of the COVID-19 case rate and the unemployment rate.

Based on above data exploratory result, the following equation is used to create a regression model
to determine the relationship between the case rate and demographic features including the seniors rate, the rate of poverty and the rate of unemployment.

@@

Accordingt to the assessment of homoskedasticity of model 2, which would be discussed in the following paragraph (Model 2 limits - iv.Homoskedastic Errors), it is plausible to apply the classical standard errors in the t-test of coefficients here.

@@

Based on the test results of the estimated coefficients in model 2, all three demographic features discussed here are significantly realated to the COVID-19 care rate, especially for the poverty rate (p-value < 0.01).

The coefficient of the senior_per_100k is  -0.230970 and statistically significant, which implies that for every one thousand additional seniors per 100,000, there is a decrease of 231 COVID-19 cases per 100,000 holding other demographic features as constant. As discussed in mode1 1 section, although seniors are susceptible group, less mobility and social interaction with cautious self procetion may play a important role in this pandemic situation.

In addition, the coefficient of the poverty_per_100k (0.231750) indicats that for every one thousand additional people living under poverty line per 100,000, there is a increase of 232 COVID-19 case per 100,000 holding other demographic features as constant. People living under the poverty line have limited accesss to medical resourses. Moreover, most of them could only afford to live in the communites with poor hygiene.  condition.

Finally, the coefficient of the unemployed_per_100k is statiscally significant and negative(-0.486191), which implies that for every one thousand additional unemployed, there is a decrease of 486 COVID-19 case per 100,000 holding other demographic features as constant. Although the unemployment leads to less medical insurace coverage and impact on the incoming, the unempolyed people may have more flexibility in obeying stay-at-home. In addition, there is less expourse probability when no commuting or in-place working is required for the unemployed.

#### Model 2 limits

##### i. IID Sampling 
(Would be discussed in the following section - 3. Limitations of the Models)

##### ii.Linear Conditional Expectation
To assess the linear conditional expectation assumption in higher-dimensional space, the plots of predicted versus residuals of the model 2 is investigated as follow.

@@

#####iii.No Perfect Collinearity

@@

#####iv.Homoskedastic Errors
To evaluate the homoskedastic errors assumption, the square root of the residuals is plotted against the fitted values of the model 2 (scale-location plot). Although, the curve is not perfect flat smooth, there is no obvious variance of the errors either. For reference, the data points marked in the plot represent Florida(10), New York (33) and North Dakota (35).

@@

#####v.Normally Distributed Errors
To access the assumption of normality of error distribution, both the histogram and the qqplot of the residuals is shown as follow.

```{r, echo=FALSE, warning=FALSE, message=False}
plot_one <- df %>% 
  ggplot(aes(x = model2_resids)) +
  geom_histogram(fill = 'skyblue', color = 'grey30', bins=30) +
  labs(title =  "Histogram of the Model 2 Residuals", 
       x = "Model 2 residuals", y = 'Count')

@@

##### Cook Distance

@@

### Model 3

@@

#### Model 3 Regression

>>>>

```{r variables, warning=FALSE, results='hide'}
df <- read.csv("covid-19.csv", header = TRUE)
df<-df%>%
  rename(case_rate_100k = 'Case.Rate.per.100000',
         population_density = 'Population.density.per.square.miles',
         mask_public='Mandate.face.mask.use.by.all.individuals.in.public.spaces',
         poverty_pct = "Percent.living.under.the.federal.poverty.line..2018.",
         unemployed_pct = "Percent.Unemployed..2018.",
         senior_pct = "X65.",
         population_2018 = "Population.2018" ## extra variable in test
         ) %>%
  select(State, case_rate_100k, population_density, mask_public, poverty_pct, 
         unemployed_pct, senior_pct, population_2018)
# Assign correct data types
head(df)
```

# 1. An Introduction

@@

## Assumptions

@@

# 2. A Model Building Process

@@

## Model Variables

@@

In order to allow for more simplistic and direct interpretations when analyzing the effect of the input variables on the output, it was determined that all percentage variables (senior, poverty, and unemployment) would be transformed into rates per 100,000 to align with the case rate variable. It will also serve to ease the burden on the readers of this report as like-for-like scaled relationships can be drawn between the variables. Given that this transformation is constant for all values (scalar multiplication of 100,000 for the seniors variable given it is a ratio, and 1,000 for poverty and unemployment given they are percentages on a scale of 0-100%), it will have no effect on the distributions of the variables.

```{r}  
# Percentage variables are re-scaled to rates per 100k
df <-df %>%
  mutate(    
    senior_per_100k = 100000*senior_pct,
    poverty_per_100k = 1000*poverty_pct, 
    unemployed_per_100k = 1000*unemployed_pct
  )
```

## Model 1

@@

### Model 1 Regression

@@

### Model 1 Limitations

#### i. IID Sampling

@@

#### ii. Linear Conditional Expectation

@@

#### iii. No Perfect Collinearity

@@

#### iv. Homoskedastic Errors

@@

#### v. Normally Distributed Errors

@@

#### vi. Influence of Data Points

@@

## Model 2

Based on the work of model 1, two extra state demographic features regarding poverty rate (percentage living under the federal poverty line in 2018) and unemployment rate (percentage of unemployed in 2018) would be included in the analysis of model 2. In addition, both newly-introduced percentage variables would be transformed into rates per 100,000 to align with the case rate variable.

@@

```{r}
ggplot(data = df,
  mapping = aes(x= poverty_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(6000,21200,800)) +
  labs(title =  "Histogram of Poverty Rate", 
       x = "Number of living under poverty line per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(6000, 21200, 1600))
```
As shown in the above distribution, there is a fairly normal distribution in the poverty number per 100,000 across the states. Additionally, there is no outlier observed within the distribution, given that the range of rates fall between 7,600 and 19,700. Therefore, there is no further transformation in need for this variable in the following analysis process for both model 2 and model 3.

@@

```{r}
ggplot(data = df,
  mapping = aes(x= unemployed_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(2000,8000,500)) +
  labs(title =  "Histogram of Unemployment Rate", 
       x = "Number of unemployed per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(2000, 8000, 1000))
```
According to the histogram above, the distribution of unemployed number per 100,000 is fairly normal distribution. Overall, it is not highly skewed or heavily tailed. Therefore, there is no further transformation needed for this variable in the following analysis process for both model 2 and model 3.

### Model 2 Regression

@@

```{r message=FALSE}
case_poverty <- df %>% 
  ggplot(aes(poverty_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = ' Association between Covid-19 case rate and the poverty rate', 
    x = 'Number of living under poverty line per 100,000',
    y = 'Covid-19 case rate per 100,000', 
    color = 'Senior num per 100k'
  )

case_unemployed <- df %>% 
  ggplot(aes(unemployed_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Association between Covid-19 case rate and the unemployment rate', 
    x = 'Number of unemployed per 100,000', 
    y = 'Covid-19 case rate per 100,000',
    color = 'Senior num per 100k'
  )

case_poverty / case_unemployed
```
However, there is an overall increasing trend of COVID-19 case rate with increasing poverty rate, while there is an inverse relationship between the variables of the COVID-19 case rate and the unemployment rate.

Based on the above data exploratory result, the following equation is used to create a regression model to determine the relationship between the case rate and demographic features including the seniors rate, the rate of poverty and the rate of unemployment.

@@

According to the assessment of homoskedasticity of model 2, which would be discussed in the following paragraph (Model 2 Limitations - iv.Homoskedastic Errors), it is plausible to apply the classical standard errors in the t-test of coefficients here.

@@

Based on the test results of the estimated coefficients in model 2, all three demographic features discussed here are significantly related to the COVID-19 case rate, especially for the poverty rate (p-value < 0.01).

The coefficient of the senior_per_100k is -0.230970 and statistically significant, which implies that for every one thousand additional seniors per 100,000, there is a decrease of 231 COVID-19 cases per 100,000, holding other demographic features as constant. As discussed in mode1 1 section, although seniors are susceptible group, less mobility and social interaction with cautious self protection may play an important role in this pandemic situation.

In addition, the coefficient of the poverty_per_100k (0.231750) indicates that for every one thousand additional people living under poverty line per 100,000, there is a increase of 232 COVID-19 cases per 100,000, holding other demographic features as constant. People living under the poverty line have limited access to medical resources. Moreover, most of them could only afford to live in the communities with poor hygiene conditions.

Finally, the coefficient of the unemployed_per_100k is statistically significant and negative (-0.486191), which implies that for every one thousand additional unemployed, there is a decrease of 486 COVID-19 case per 100,000, holding other demographic features as constant. Although the unemployment leads to less medical insurance coverage and impact on their income, the unemployed people may have more flexibility to obey stay-at-home orders. In addition, there is less exposure probability when there is no commuting or in-place working required for the unemployed.

### Model 2 Limitations

#### i. IID Sampling 
(Would be discussed in the following section - 3. Limitations of the Models)

#### ii.Linear Conditional Expectation
To assess the linear conditional expectation assumption in higher-dimensional space, the plots of predicted versus residuals of the model 2 is investigated as follow.

@@

#### iii.No Perfect Collinearity

@@

#### iv.Homoskedastic Errors
To evaluate the homoskedastic errors assumption, the square root of the residuals is plotted against the fitted values of the model 2 (scale-location plot). Although, the curve is not perfect flat smooth, there is no obvious variance of the errors either. For reference, the data points marked in the plot represent Florida(10), New York (33) and North Dakota (35).

@@

#### v.Normally Distributed Errors
To access the assumption of normality of error distribution, both the histogram and the qqplot of the residuals is shown as follow.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
plot_one <- df %>% 
  ggplot(aes(x = model2_resids)) +
  geom_histogram(fill = 'skyblue', color = 'grey30', bins=30) +
  labs(title =  "Histogram of the Model 2 Residuals", 
       x = "Model 2 residuals", y = 'Count')

@@

### Cook Distance

@@

## Model 3

@@

### Model 3 Regression
